---
title: "Chemistry calculations"
vignette: >
  %\VignetteIndexEntry{Chemistry calculations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Stream metabolism workflows often need dissolved gas calculations before model
fitting or quality checks. preMetabolizer includes helpers for oxygen
saturation, CO2 partial pressure, dissolved CO2 concentration, water vapor
pressure, and CO2 solubility.

```{r}
#| label: libraries
#| message: false

library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Dissolved oxygen saturation

`calc_O2sat()` calculates dissolved oxygen saturation in mg/L from water
temperature, barometric pressure, pressure units, and optional salinity. This is
the `DO.sat` input commonly used by stream metabolism models.

```{r}
#| label: oxygen-saturation

water <- tibble::tibble(
  temp_water = seq(5, 30, by = 5),
  atmo_press = 101.325,
  salinity = 0
) |>
  mutate(
    DO_sat_mg_L = calc_O2sat(
      temp_water = temp_water,
      atmo_press = atmo_press,
      units = "kPa",
      salinity = salinity
    )
  )

water
```

Oxygen saturation declines as water warms.

```{r}
#| label: plot-oxygen-saturation

ggplot(water, aes(temp_water, DO_sat_mg_L)) +
  geom_line(color = "#2c7fb8") +
  geom_point(color = "#2c7fb8") +
  labs(
    x = "Water temperature (C)",
    y = "O2 saturation (mg/L)"
  ) +
  theme_bw()
```

## CO2 partial pressure

`xCO2_to_pCO2()` converts CO2 mole fraction in air to water-vapor-corrected
partial pressure. `pCO2_to_xCO2()` performs the inverse conversion.

```{r}
#| label: co2-partial-pressure

pCO2 <- xCO2_to_pCO2(
  xCO2_ppm = c(420, 800, 1200),
  temp_water = 20,
  atmo_press = 101.325,
  press_units = "kPa",
  salinity = 0,
  method = "MIMSY"
)

pCO2

pCO2_to_xCO2(
  temp_water = 20,
  pCO2_uatm = pCO2,
  atmo_press = 101.325,
  press_units = "kPa",
  salinity = 0,
  method = "MIMSY"
)
```

## Dissolved CO2 concentration

`calc_CO2_molKg()` and `calc_CO2_mgL()` estimate dissolved CO2 concentration
from CO2 mole fraction, water temperature, water depth, atmospheric pressure,
and salinity.

```{r}
#| label: co2-concentration

co2 <- tibble::tibble(
  CO2_ppm = c(420, 800, 1200),
  temp_water = 20,
  water_depth_m = 0.5,
  atmo_press = 101.325
) |>
  mutate(
    CO2_mol_kg = calc_CO2_molKg(
      CO2_ppm = CO2_ppm,
      temp_water = temp_water,
      water_depth_m = water_depth_m,
      atmo_press = atmo_press,
      press_units = "kPa"
    ),
    CO2_mg_L = calc_CO2_mgL(
      CO2_ppm = CO2_ppm,
      temp_water = temp_water,
      water_depth_m = water_depth_m,
      atmo_press = atmo_press,
      press_units = "kPa"
    )
  )

co2
```

## Vapor pressure and solubility

`calc_vapor_press()` returns water vapor pressure in atmospheres. Use
`method = "MIMSY"` for freshwater calculations and `"Dickson2007"` for
seawater-style salinity corrections. `calc_K0()` returns the Weiss (1974) CO2
solubility coefficient.

```{r}
#| label: vapor-pressure-k0

temps <- tibble::tibble(temp_water = c(5, 15, 25)) |>
  mutate(
    vapor_press_atm = calc_vapor_press(
      temp_water,
      salinity = 0,
      method = "MIMSY"
    ),
    K0_mol_kg_atm = calc_K0(temp_water, water_depth_m = 0.5)
  )

temps
```

## Combine calculations for model input

The calculations are vectorized, so they fit naturally into a data preparation
pipeline.

```{r}
#| label: model-input

logger <- tibble::tibble(
  dateTime = as.POSIXct(
    c("2024-06-01 00:00:00", "2024-06-01 01:00:00"),
    tz = "UTC"
  ),
  temp_water = c(18.5, 18.2),
  atmo_kPa = c(99.8, 99.7),
  xCO2_ppm = c(600, 650)
) |>
  mutate(
    DO.sat = calc_O2sat(temp_water, atmo_kPa, units = "kPa"),
    pCO2_uatm = xCO2_to_pCO2(
      xCO2_ppm = xCO2_ppm,
      temp_water = temp_water,
      atmo_press = atmo_kPa,
      press_units = "kPa",
      salinity = 0,
      method = "MIMSY"
    )
  )

logger
```
