---
title: "Preparing French Creek Data for Stream Metabolism Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing French Creek Data for Stream Metabolism Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r libraries, include=FALSE}
library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Introduction

Stream metabolism models estimate gross primary production and ecosystem
respiration from continuous sensor records. The
[streamMetabolizer](https://github.com/ConnorB/streamMetabolizer) package fits
these models, but it requires a specific set of input variables:

| Column | Description |
|---|---|
| `solar.time` | Mean solar time (POSIXct, tz = "UTC") |
| `DO.obs` | Observed dissolved oxygen (mg/L) |
| `DO.sat` | DO saturation at current conditions (mg/L) |
| `depth` | Mean stream depth (m) |
| `temp.water` | Water temperature (°C) |
| `light` | Photosynthetically active radiation (µmol/m²/s) |

This vignette walks through computing each of those inputs from the
`french_creek` dataset using preMetabolizer. The data come from a field study
on French Creek near Laramie, Wyoming, USA (Hall et al. 2016,
<https://doi.org/10.1890/14-0631.1>).

## The `french_creek` dataset

```{r load}
data(french_creek)
french_creek <- french_creek |> 
  filter(lubridate::minute(datetime) %in% c(0, 15, 30, 45))

glimpse(french_creek)
```


The dataset covers 15-minute intervals from August 23 to September 30, 2012.
Two sensors were deployed at the site: `"REZN"` ran from the start of the study
through September 1 at 12:50 MDT, when it was replaced by `"TOWN"`. The plot
below shows the full record for both sensors.

```{r plot-raw, fig.width = 7, fig.height = 5, fig.alt = "Time series of water temperature and dissolved oxygen at French Creek, colored by sensor (REZN and TOWN)."}
french_creek |>
  tidyr::pivot_longer(
    c(temp_C, DO_mgL),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(variable = recode(
    variable,
    temp_C = "Temperature (°C)",
    DO_mgL = "DO (mg/L)"
  )) |>
  ggplot(aes(datetime, value, color = sonde)) +
  geom_line(linewidth = 0.3, alpha = 0.8) +
  facet_wrap(~variable, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c(REZN = "#E69F00", TOWN = "#0072B2")) +
  labs(x = NULL, y = NULL, color = "Sensor") +
  theme_bw()
```

The REZN sensor shows a cluster of anomalous negative temperature values near
the end of its deployment. For the rest of this vignette we filter to the TOWN
sensor, which has the longer and cleaner record.

```{r filter-town}
french_creek <- french_creek |>
  filter(sonde == "TOWN") |>
  select(-sonde)
```

## Remove anomalous values

Negative water temperatures are physically impossible for liquid water and
indicate sensor malfunction. DO values <= 5 mg/L are improbable and may indicate sensor malfunction. We'll remove those rows.

```{r remove-anomalous}
french_creek <- french_creek |>
  filter(is.na(temp_C) | temp_C >= 0) |> 
  filter(is.na(DO_mgL) | DO_mgL > 5)
```

## Ensure even timesteps

streamMetabolizer requires evenly spaced data. `even_timesteps()`
fills any gaps with `NA` rows so the interval is consistent.

```{r even-timesteps}
french_creek <- even_timesteps(french_creek, datetime_col = "datetime")
```


## Convert datetime to solar time

streamMetabolizer works in mean solar time, which shifts UTC by the
longitude-based offset so that noon falls roughly when the sun is highest.
We define site constants once and reuse them throughout.

```{r solar-time}
lat <- 41.33
lon <- -106.3

french_creek <- french_creek |>
  mutate(
    solar.time = convert_UTC_to_solartime(
      datetime,
      longitude = lon,
      time.type = "mean solar"
    )
  ) |> 
  filter(solar.time >= as.POSIXct("2012-09-18 04:05:58") &
         solar.time <= as.POSIXct("2012-09-21 03:50:58"))
```

## Calculate modeled light (PAR)

`calc_light()` returns photosynthetically active radiation in µmol/m²/s based
on the solar geometry at the site. It takes mean solar time as input.

```{r light}
french_creek <- french_creek |>
  mutate(light = calc_light(solar.time, latitude = lat, longitude = lon))
```

## Adjust barometric pressure for elevation

French Creek sits at roughly 2,195 m (7,200 ft) above sea level. Using
standard sea-level pressure (101.325 kPa) would overpredict O~2~ saturation
by about 20%, so we correct it down to site elevation using `correct_bp()`.

```{r bp}
site_bp_kPa <- correct_bp(
  station_bp   = 101.325,
  air_temp     = 10,
  station_elev = 0,
  site_elev    = 2195,
  from_units   = "kPa",
  to_units     = "kPa"
)

site_bp_kPa

#streamMetab calc
#site_bp_kPa <- convert_pressure(pressure = 523 * 1.33322368,from = "mbar", to = "kPa")# 523 mmHg * 1.33322368 mb mmHg^-1 -> 697.27 mb, applicable at ~10000 ft

#site_bp_kPa
```

In a real study, replace the scalar 101.325 kPa with a time series of observed
barometric pressure from a nearby weather station or on-site logger.

## Calculate O~2~ saturation

With site-adjusted pressure and water temperature, `calc_O2sat()` returns the
DO saturation concentration in mg/L.

```{r o2sat}
french_creek <- french_creek |>
  mutate(
    DO.sat = calc_O2sat(
      temp_water = temp_C,
      atmo_press = site_bp_kPa,
      units      = "kPa"
    )
  )

summary(french_creek$DO.sat)
```

## Assemble the streamMetabolizer input

The final step restructures the data into the columns streamMetabolizer
expects.

```{r sm-input}
sm_input <- french_creek |>
  transmute(
    solar.time = solar.time,
    DO.obs     = DO_mgL,
    DO.sat     = DO.sat,
    depth      = 0.16,
    temp.water = temp_C,
    light      = light
  )

head(sm_input)
```

```{r plot-sm-input-do, fig.width = 7, fig.height = 5, fig.alt = "Time series of observed DO, saturated DO, and percent DO saturation at French Creek."}
sm_input |>
  mutate(DO.pctsat = 100 * DO.obs / DO.sat) |>
  tidyr::pivot_longer(
    c(DO.obs, DO.sat, DO.pctsat),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(units = ifelse(variable == "DO.pctsat", "DO\n(% sat)", "DO\n(mg/L)")) |>
  ggplot(aes(solar.time, value, color = variable)) +
  geom_line(linewidth = 0.3) +
  facet_grid(units ~ ., scales = "free_y") +
  labs(x = "Solar time", y = NULL, color = "Variable") +
  theme_bw()
```

```{r plot-sm-input-other, fig.width = 7, fig.height = 6, fig.alt = "Time series of depth, water temperature, and PAR at French Creek."}
labels <- c(
  depth      = "depth~(m)",
  temp.water = "water~temp~(degree*C)",
  light      = "PAR~(mu*mol~m^{-2}~s^{-1})"
)

sm_input |>
  tidyr::pivot_longer(
    c(depth, temp.water, light),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(
    variable = ordered(variable, levels = c("depth", "temp.water", "light"))
  ) |>
  ggplot(aes(solar.time, value, color = variable)) +
  geom_line(linewidth = 0.3) +
  facet_grid(
    variable ~ .,
    scales = "free_y",
    labeller = ggplot2::as_labeller(labels, label_parsed)
  ) +
  labs(x = "Solar time", y = NULL, color = "Variable") +
  theme_bw()
```

This tibble is ready to pass to `streamMetabolizer::metab()`. See the
[streamMetabolizer documentation](https://github.com/ConnorB/streamMetabolizer)
for details on model fitting and interpreting metabolism estimates.

## References

Hall, R. O., Jr., Tank, J. L., Baker, M. A., Rosi-Marshall, E. J., &
Hotchkiss, E. R. (2016). Metabolism, gas exchange, and carbon spiraling in
rivers. *Ecosystems*, 19(1), 73--86. <https://doi.org/10.1890/14-0631.1>
