---
title: "Preparing French Creek Data for Stream Metabolism Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing French Creek Data for Stream Metabolism Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r libraries, include=FALSE}
library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Introduction

Stream metabolism models estimate gross primary production and ecosystem
respiration from continuous sensor records. The
[streamMetabolizer](https://github.com/ConnorB/streamMetabolizer) package fits
these models, but it requires a specific set of input variables:

| Column | Description |
|---|---|
| `solar.time` | Mean solar time (POSIXct, tz = "UTC") |
| `DO.obs` | Observed dissolved oxygen (mg/L) |
| `DO.sat` | DO saturation at current conditions (mg/L) |
| `depth` | Mean stream depth (m) |
| `temp.water` | Water temperature (°C) |
| `light` | Photosynthetically active radiation (µmol/m²/s) |

This vignette walks through computing each of those inputs from the
`french_creek` dataset using preMetabolizer. The data come from a field study
on French Creek near Laramie, Wyoming, USA (Hall et al. 2016,
<https://doi.org/10.1890/14-0631.1>).
Below we keep NASA-derived PAR as `light.obs` and modeled PAR from
`calc_light()` as `light.calc` so the two light series can be compared.

## The `french_creek` dataset

```{r load}
data(french_creek)
french_creek <- french_creek |> 
  filter(lubridate::minute(datetime) %in% c(0, 15, 30, 45))

glimpse(french_creek)
```


The dataset covers 15-minute intervals from August 23 to September 30, 2012.
Two sensors were deployed at the site: `"REZN"` ran from the start of the study
through September 1 at 12:50 MDT, when it was replaced by `"TOWN"`. The plot
below shows the full record for both sensors.

```{r plot-raw, fig.width = 7, fig.height = 5, fig.alt = "Time series of water temperature and dissolved oxygen at French Creek, colored by sensor (REZN and TOWN)."}
french_creek |>
  tidyr::pivot_longer(
    c(temp_C, DO_mgL),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(variable = recode(
    variable,
    temp_C = "Temperature (°C)",
    DO_mgL = "DO (mg/L)"
  )) |>
  ggplot(aes(datetime, value, color = sonde)) +
  geom_line(linewidth = 0.3, alpha = 0.8) +
  facet_wrap(~variable, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c(REZN = "#E69F00", TOWN = "#0072B2")) +
  labs(x = NULL, y = NULL, color = "Sensor") +
  theme_bw()
```

The REZN sensor shows a cluster of anomalous negative temperature values near
the end of its deployment. For the rest of this vignette we filter to the TOWN
sensor, which has the longer and cleaner record.

```{r filter-town}
french_creek <- french_creek |>
  filter(sonde == "TOWN") |>
  select(-sonde)
```

## Remove anomalous values

Negative water temperatures are physically impossible for liquid water and
indicate sensor malfunction. DO values <= 5 mg/L are improbable and may indicate sensor malfunction. We'll remove those rows.

```{r remove-anomalous}
french_creek <- french_creek |>
  filter(is.na(temp_C) | temp_C >= 0) |> 
  filter(is.na(DO_mgL) | DO_mgL > 5)
```

## Ensure even timesteps

streamMetabolizer requires evenly spaced data. `even_timesteps()`
fills any gaps with `NA` rows so the interval is consistent.

```{r even-timesteps}
french_creek <- even_timesteps(french_creek, datetime_col = "datetime")
```


## Convert datetime to solar time

streamMetabolizer works in mean solar time, which shifts UTC by the
longitude-based offset so that noon falls roughly when the sun is highest.
We define site constants once and reuse them throughout.

```{r solar-time}
lat <- 41.33
lon <- -106.3

french_creek <- french_creek |>
  mutate(
    solar.time = convert_UTC_to_solartime(
      datetime,
      longitude = lon,
      time.type = "mean solar"
    )
  ) |> 
  filter(solar.time >= as.POSIXct("2012-09-18 04:05:58") &
         solar.time <= as.POSIXct("2012-09-21 03:50:58"))
```

## Calculate modeled light (PAR)

`calc_light()` returns photosynthetically active radiation in µmol/m²/s based
on the solar geometry at the site. It takes mean solar time as input.

```{r light}
french_creek <- french_creek |>
  mutate(light.calc = calc_light(solar.time, latitude = lat, longitude = lon))
```

## Get barometric pressure and observed light

French Creek sits at roughly 2,195 m (7,200 ft) above sea level. Using
standard sea-level pressure (101.325 kPa) would overpredict O~2~ saturation by
about 20%, so we download NASA POWER surface pressure for the site with
`get_nasa_data()`. We also download shortwave irradiance; `get_nasa_data()`
converts it to observed PAR as `light.obs` and interpolates NASA fields to the
same timestamps as `french_creek`.

```{r bp}
french_elev_m <- get_usgs_elev(
  latitude = lat,
  longitude = lon,
  units = "Meters"
)

nasa_dat <- get_nasa_data(
  data = french_creek,
  datetime_col = "datetime",
  lat = lat,
  lon = lon,
  elev_m = french_elev_m,
  params = c("PSC", "ALLSKY_SFC_SW_DWN")
) |>
  select(datetime = dateTime, bp_kPa = PSC, light.obs)

nasa_dat

french_data <- french_creek |>
  left_join(nasa_dat, by = "datetime")
```

## Calculate O~2~ saturation

With site-adjusted pressure and water temperature, `calc_O2sat()` returns the
DO saturation concentration in mg/L.

```{r o2sat}
french_data <- french_data |>
  mutate(
    DO.sat = calc_O2sat(
      temp_water = temp_C,
      atmo_press = bp_kPa,
      units      = "kPa"
    )
  )

summary(french_data$DO.sat)
```

## Assemble the final tibble

The final step restructures the data and retains both observed and calculated
light.

```{r sm-input}
sm_input <- french_data |>
  transmute(
    solar.time = solar.time,
    DO.obs     = DO_mgL,
    DO.sat     = DO.sat,
    depth      = 0.16,
    temp.water = temp_C,
    light.obs  = light.obs,
    light.calc = light.calc
  )

head(sm_input)
```

```{r plot-sm-input-do, fig.width = 7, fig.height = 5, fig.alt = "Time series of observed DO, saturated DO, and percent DO saturation at French Creek."}
sm_input |>
  mutate(DO.pctsat = 100 * DO.obs / DO.sat) |>
  tidyr::pivot_longer(
    c(DO.obs, DO.sat, DO.pctsat),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(units = ifelse(variable == "DO.pctsat", "DO\n(% sat)", "DO\n(mg/L)")) |>
  ggplot(aes(solar.time, value, color = variable)) +
  geom_line(linewidth = 0.3) +
  facet_grid(units ~ ., scales = "free_y") +
  labs(x = "Solar time", y = NULL, color = "Variable") +
  theme_bw()
```

```{r plot-sm-input-other, fig.width = 7, fig.height = 6, fig.alt = "Time series of depth, water temperature, observed PAR, and calculated PAR at French Creek."}
labels <- c(
  depth      = "depth~(m)",
  temp.water = "water~temp~(degree*C)",
  light.obs  = "atop(observed~PAR, (mu*mol~m^{-2}~s^{-1}))",
  light.calc = "atop(calculated~PAR, (mu*mol~m^{-2}~s^{-1}))"
)

sm_input |>
  tidyr::pivot_longer(
    c(depth, temp.water, light.obs, light.calc),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(
    variable = ordered(
      variable,
      levels = c("depth", "temp.water", "light.obs", "light.calc")
    )
  ) |>
  ggplot(aes(solar.time, value, color = variable)) +
  geom_line(linewidth = 0.3) +
  facet_grid(
    variable ~ .,
    scales = "free_y",
    labeller = ggplot2::as_labeller(labels, label_parsed)
  ) +
  labs(x = "Solar time", y = NULL, color = "Variable") +
  theme_bw()
```

To fit a model with `streamMetabolizer::metab()`, choose either `light.obs` or
`light.calc` and rename that column to `light`. See the
[streamMetabolizer documentation](https://github.com/ConnorB/streamMetabolizer)
for details on model fitting and interpreting metabolism estimates.

## References

Hall, R. O., Jr., Tank, J. L., Baker, M. A., Rosi-Marshall, E. J., &
Hotchkiss, E. R. (2016). Metabolism, gas exchange, and carbon spiraling in
rivers. *Ecosystems*, 19(1), 73--86. <https://doi.org/10.1890/14-0631.1>
