---
title: "Working with Iowa Environmental Mesonet data"
vignette: >
  %\VignetteIndexEntry{Working with Iowa Environmental Mesonet data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

cache_dir <- tools::R_user_dir("preMetabolizer", which = "cache")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

.run <- tryCatch({
  httr2::request("https://mesonet.agron.iastate.edu") |>
    httr2::req_timeout(5) |>
    httr2::req_perform()
  TRUE
}, error = function(e) {
  length(list.files(cache_dir, pattern = "^iem_")) > 0
})

knitr::opts_chunk$set(eval = .run)
```

## Introduction

The [Iowa Environmental Mesonet](https://mesonet.agron.iastate.edu/) (IEM) provides public access to many meteorological and hydrological observing networks. IEM groups stations into network identifiers such as `IA_ASOS`, `IA_COOP`, and many state or special-purpose networks.

preMetabolizer provides helpers for selected IEM API v1 endpoints:

- `iem_networks()` lists available IEM network identifiers.
- `iem_stations()` retrieves station metadata for one network.
- `iem_station()` looks up one station identifier across networks.
- `iem_current()` retrieves current observations.
- `iem_obhistory()` retrieves one local day of station observations.
- `iem_daily()` retrieves daily summary observations.

> **Note:** These functions contact IEM web services and require an internet connection. Code chunks that call the API will not run during package installation if the service is unreachable. Run them interactively in your own session.

```{r}
#| label: libraries
#| message: false

library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Caching downloaded data

Repeated calls to the IEM API download the same data on every run. This vignette saves each result to a local cache directory the first time it is downloaded and reloads from disk on subsequent runs. The cache lives in `tools::R_user_dir("preMetabolizer", which = "cache")`, a platform-appropriate, user-specific directory that persists across sessions. Each data-fetching chunk below checks for a cached `.rds` file, downloads and saves on the first run, and reloads from disk on all subsequent runs.

## Discover networks and stations

Start with `iem_networks()` when you need to find the exact network identifier for a station source. The `network` column is the value to pass to other IEM helpers.

```{r}
#| label: networks

cache_file <- file.path(cache_dir, "iem_networks.rds")
if (!file.exists(cache_file)) {
  networks <- iem_networks()
  saveRDS(networks, cache_file)
} else {
  networks <- readRDS(cache_file)
}

networks |>
  filter(grepl("Iowa", network_name)) |>
  select(network, network_name, tzname)
```

Station metadata for a network includes station identifiers, names, coordinates, archive dates, and the station time zone when IEM provides it.

```{r}
#| label: stations

cache_file <- file.path(cache_dir, "iem_stations_ia_asos.rds")
if (!file.exists(cache_file)) {
  ia_asos <- iem_stations("IA_ASOS")
  saveRDS(ia_asos, cache_file)
} else {
  ia_asos <- readRDS(cache_file)
}

ia_asos |>
  select(station_id, station_name, network, tzname, archive_begin, longitude, latitude) |>
  arrange(station_id)
```

Some station identifiers appear in more than one network. Use `iem_station()` when you want to check how IEM resolves a station ID before requesting data.

```{r}
#| label: station-lookup

cache_file <- file.path(cache_dir, "iem_station_amw.rds")
if (!file.exists(cache_file)) {
  amw <- iem_station("AMW")
  saveRDS(amw, cache_file)
} else {
  amw <- readRDS(cache_file)
}

amw |>
  select(station_id, station_name, network, tzname, archive_begin)
```

## Retrieve current observations

`iem_current()` requires at least one filter. You can request all current observations in a network, stations within a network, stations by forecast office, or one or more station identifiers.

```{r}
#| label: current-network

cache_file <- file.path(cache_dir, "iem_current_ia_asos.rds")
if (!file.exists(cache_file)) {
  current <- iem_current(network = "IA_ASOS")
  saveRDS(current, cache_file)
} else {
  current <- readRDS(cache_file)
}

current |>
  select(station_id, station_name, utc_valid, tmpf, dwpf, sknt, drct, mslp) |>
  arrange(station_id)
```

For station-specific work, include both `network` and `stations` so shared station identifiers do not surprise you.

```{r}
#| label: current-stations

cache_file <- file.path(cache_dir, "iem_current_dsm_amw.rds")
if (!file.exists(cache_file)) {
  dsm_current <- iem_current(
    network = "IA_ASOS",
    stations = c("DSM", "AMW"),
    minutes = 120
  )
  saveRDS(dsm_current, cache_file)
} else {
  dsm_current <- readRDS(cache_file)
}

dsm_current |>
  select(station_id, station_name, utc_valid, tmpf, relh, sknt, pres)
```

## Retrieve one day of observations

`iem_obhistory()` returns observations for a single station and local station date. UTC timestamp columns are parsed as UTC `POSIXct` values. Local timestamp columns are returned as character values because station time zones vary across networks.

```{r}
#| label: obhistory

cache_file <- file.path(cache_dir, "iem_obhistory_dsm_2024-06-01.rds")
if (!file.exists(cache_file)) {
  obs <- iem_obhistory(
    station = "DSM",
    network = "IA_ASOS",
    date = "2024-06-01"
  )
  saveRDS(obs, cache_file)
} else {
  obs <- readRDS(cache_file)
}

obs |>
  select(utc_valid, local_valid, tmpf, dwpf, sknt, drct, p01_i)
```

Hourly or sub-hourly records can be plotted directly after selecting the variables you need.

```{r}
#| label: obhistory-plot

obs |>
  filter(!is.na(tmpf)) |>
  ggplot(aes(utc_valid, tmpf)) +
  geom_line() +
  labs(
    x = "UTC time",
    y = "Air temperature (deg F)"
  )
```

## Retrieve daily summaries

Use `iem_daily()` for daily summary observations. Provide either one `date`, or provide `year` and optional `month` for a longer request.

```{r}
#| label: daily-station

cache_file <- file.path(cache_dir, "iem_daily_dsm_2024-06.rds")
if (!file.exists(cache_file)) {
  dsm_daily <- iem_daily(
    network = "IA_ASOS",
    station = "DSM",
    year = 2024,
    month = 6
  )
  saveRDS(dsm_daily, cache_file)
} else {
  dsm_daily <- readRDS(cache_file)
}

dsm_daily |>
  select(station_id, date, max_tmpf, min_tmpf, precip) |>
  arrange(date)
```

For a one-day network request, omit `station` and supply `date`.

```{r}
#| label: daily-network

cache_file <- file.path(cache_dir, "iem_daily_ia_asos_2024-06-01.rds")
if (!file.exists(cache_file)) {
  ia_day <- iem_daily("IA_ASOS", date = "2024-06-01")
  saveRDS(ia_day, cache_file)
} else {
  ia_day <- readRDS(cache_file)
}

ia_day |>
  select(station_id, date, max_tmpf, min_tmpf, precip) |>
  arrange(desc(date))
```

## Practical notes

IEM combines many observing networks, and each network can have different available variables, units, reporting intervals, and station identifiers. A good workflow is to first inspect `iem_networks()` and `iem_stations()`, then request observations from a specific `network` and `station` pair.
