---
title: "Working with Iowa Environmental Mesonet data"
vignette: >
  %\VignetteIndexEntry{Working with Iowa Environmental Mesonet data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The [Iowa Environmental Mesonet](https://mesonet.agron.iastate.edu/) (IEM) provides public access to many meteorological and hydrological observing networks. IEM groups stations into network identifiers such as `IA_ASOS`, `IA_COOP`, and many state or special-purpose networks.

preMetabolizer provides helpers for selected IEM API v1 endpoints:

- `iem_networks()` lists available IEM network identifiers.
- `iem_stations()` retrieves station metadata for one network.
- `iem_station()` looks up one station identifier across networks.
- `iem_current()` retrieves current observations.
- `iem_obhistory()` retrieves one local day of station observations.
- `iem_daily()` retrieves daily summary observations.

> **Note:** These functions contact IEM web services and require an internet connection. Code chunks marked with `#| eval: false` will not run during package installation. Run them interactively in your own session.

```{r}
#| label: libraries
#| message: false

library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Discover networks and stations

Start with `iem_networks()` when you need to find the exact network identifier for a station source. The `id` column is the value to pass to other IEM helpers.

```{r}
#| label: networks
#| eval: false

networks <- iem_networks()

networks |>
  filter(grepl("Iowa", name)) |>
  select(id, name, tzname)
```

Station metadata for a network includes station identifiers, names, coordinates, archive dates, and the station time zone when IEM provides it.

```{r}
#| label: stations
#| eval: false

ia_asos <- iem_stations("IA_ASOS")

ia_asos |>
  select(id, name, network, tzname, archive_begin, longitude, latitude) |>
  arrange(id)
```

Some station identifiers appear in more than one network. Use `iem_station()` when you want to check how IEM resolves a station ID before requesting data.

```{r}
#| label: station-lookup
#| eval: false

iem_station("AMW") |>
  select(id, name, network, tzname, archive_begin)
```

## Retrieve current observations

`iem_current()` requires at least one filter. You can request all current observations in a network, stations within a network, stations by forecast office, or one or more station identifiers.

```{r}
#| label: current-network
#| eval: false

current <- iem_current(network = "IA_ASOS")

current |>
  select(station, name, utc_valid, tmpf, dwpf, sknt, drct, mslp) |>
  arrange(station)
```

For station-specific work, include both `network` and `stations` so shared station identifiers do not surprise you.

```{r}
#| label: current-stations
#| eval: false

dsm_current <- iem_current(
  network = "IA_ASOS",
  stations = c("DSM", "AMW"),
  minutes = 120
)

dsm_current |>
  select(station, name, utc_valid, tmpf, relh, sknt, p01i)
```

## Retrieve one day of observations

`iem_obhistory()` returns observations for a single station and local station date. UTC timestamp columns are parsed as UTC `POSIXct` values. Local timestamp columns are returned as character values because station time zones vary across networks.

```{r}
#| label: obhistory
#| eval: false

obs <- iem_obhistory(
  station = "DSM",
  network = "IA_ASOS",
  date = "2024-06-01"
)

obs |>
  select(utc_valid, local_valid, tmpf, dwpf, sknt, drct, p01i)
```

Hourly or sub-hourly records can be plotted directly after selecting the variables you need.

```{r}
#| label: obhistory-plot
#| eval: false

obs |>
  filter(!is.na(tmpf)) |>
  ggplot(aes(utc_valid, tmpf)) +
  geom_line() +
  labs(
    x = "UTC time",
    y = "Air temperature (deg F)"
  )
```

## Retrieve daily summaries

Use `iem_daily()` for daily summary observations. Provide either one `date`, or provide `year` and optional `month` for a longer request.

```{r}
#| label: daily-station
#| eval: false

dsm_daily <- iem_daily(
  network = "IA_ASOS",
  station = "DSM",
  year = 2024,
  month = 6
)

dsm_daily |>
  select(station, day, high, low, precip) |>
  arrange(day)
```

For a one-day network request, omit `station` and supply `date`.

```{r}
#| label: daily-network
#| eval: false

ia_day <- iem_daily("IA_ASOS", date = "2024-06-01")

ia_day |>
  select(station, day, high, low, precip) |>
  arrange(desc(precip))
```

## Practical notes

IEM combines many observing networks, and each network can have different available variables, units, reporting intervals, and station identifiers. A good workflow is to first inspect `iem_networks()` and `iem_stations()`, then request observations from a specific `network` and `station` pair.
