---
title: "Downloading Kansas Mesonet data"
vignette: >
  %\VignetteIndexEntry{Downloading Kansas Mesonet data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

cache_dir <- tools::R_user_dir("preMetabolizer", which = "cache")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

.run <- tryCatch({
  httr2::request("https://mesonet.k-state.edu") |>
    httr2::req_timeout(5) |>
    httr2::req_perform()
  TRUE
}, error = function(e) {
  length(list.files(cache_dir, pattern = "^ks_meso_")) > 0
})

knitr::opts_chunk$set(eval = .run)
```

## Introduction

The [Kansas Mesonet](https://mesonet.k-state.edu/) is a statewide network of automated weather stations operated by Kansas State University. It provides sub-hourly meteorological data — including air temperature, relative humidity, precipitation, solar radiation, wind, and soil measurements — for more than 60 stations across Kansas.

This vignette walks through the workflow for discovering stations, browsing the variable catalog, downloading data, and reading cached files using the preMetabolizer Mesonet functions. We use **Konza Prairie**, a long-term ecological research (LTER) site in northeastern Kansas, as our example station.

> **Note:** Functions that contact the Mesonet API require an internet connection. Code chunks that call the API will not run during package installation if the service is unreachable. Run them interactively in your own session.

```{r}
#| label: libraries
#| message: false

library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Caching downloaded data

Repeated calls to the Kansas Mesonet API download the same data on every run. This vignette saves each result to a local cache directory the first time it is downloaded and reloads from disk on subsequent runs. The cache lives in `tools::R_user_dir("preMetabolizer", which = "cache")`, a platform-appropriate, user-specific directory that persists across sessions. Each data-fetching chunk below checks for a cached `.rds` file, downloads and saves on the first run, and reloads from disk on all subsequent runs.

## Explore the variable catalog

`ks_meso_vars()` retrieves and parses the current Kansas Mesonet variable metadata, including CSV headings, variable names, units, and descriptions.

```{r}
#| label: var-catalog

cache_file <- file.path(cache_dir, "ks_meso_vars.rds")
if (!file.exists(cache_file)) {
  mesonet_vars <- ks_meso_vars()
  saveRDS(mesonet_vars, cache_file)
} else {
  mesonet_vars <- readRDS(cache_file)
}
```

Search the catalog to find variables of interest:

```{r}
#| label: search-vars

mesonet_vars |>
  filter(grepl("temp|precip|solar|humidity", desc, ignore.case = TRUE)) |>
  select(var, tidy_name, units, desc)
```

## Discover stations

`ks_meso_stations()` retrieves metadata for all Mesonet stations, including location, network affiliation, and whether the station supports FW13 fire weather reports.

```{r}
#| label: stations

cache_file <- file.path(cache_dir, "ks_meso_stations.rds")
if (!file.exists(cache_file)) {
  stations <- ks_meso_stations()
  saveRDS(stations, cache_file)
} else {
  stations <- readRDS(cache_file)
}

glimpse(stations)
```

Filter to find Konza Prairie and confirm the exact station name used by the API:

```{r}
#| label: get-konza

konza <- stations |>
  filter(grepl("Konza", station_name, ignore.case = TRUE))

konza
```

The station name recognized by the API is `"Konza Prairie"`.

## Check station activity

`ks_meso_station_activity()` shows the observation intervals available at each station and the date range of archived data. This is useful for verifying that a station has data for your study period and identifying the finest available temporal resolution.

```{r}
#| label: check-activity

cache_file <- file.path(cache_dir, "ks_meso_station_activity.rds")
if (!file.exists(cache_file)) {
  activity <- ks_meso_station_activity()
  saveRDS(activity, cache_file)
} else {
  activity <- readRDS(cache_file)
}

activity |>
  filter(station_name == "Konza Prairie")
```

## Check the most recent observation

`ks_meso_most_recent()` returns the timestamp of the latest ingested observation for every station at a given interval, which is helpful for monitoring live data pipelines.

```{r}
#| label: check-most-recent

cache_file <- file.path(cache_dir, "ks_meso_most_recent_hour.rds")
if (!file.exists(cache_file)) {
  recent <- ks_meso_most_recent(interval = "hour")
  saveRDS(recent, cache_file)
} else {
  recent <- readRDS(cache_file)
}

recent |>
  filter(station_name == "Konza Prairie")
```

## Retrieve time-series data

`ks_meso_time_series()` retrieves Mesonet data for one or more stations and returns a tibble directly. Large date ranges are split into chunks automatically to stay within the API record limit.

Here we retrieve hourly data for Konza Prairie for the 2024 calendar year, requesting air temperature, relative humidity, precipitation, and surface pressure.

```{r}
#| label: timeseries-data

cache_file <- file.path(cache_dir, "ks_meso_konza_hourly_2024.rds")
if (!file.exists(cache_file)) {
  konza_hourly <- ks_meso_time_series(
    stations   = "Konza Prairie",
    start_date = "2024-01-01",
    end_date   = "2024-12-31",
    interval   = "hour",
    vars       = c("TEMP2MAVG", "RELHUM2MAVG", "PRECIP", "PRESSUREAVG")
  )
  saveRDS(konza_hourly, cache_file)
} else {
  konza_hourly <- readRDS(cache_file)
}

glimpse(konza_hourly)
```

## Example workflow: monthly precipitation

With the data in hand, standard `dplyr` and `ggplot2` workflows apply. The example below summarizes monthly precipitation totals and daily temperature ranges.

```{r}
#| label: monthly-ppt

monthly_precip <- konza_hourly |>
  mutate(month = lubridate::floor_date(timestamp, "month")) |>
  group_by(month) |>
  summarise(
    precip_mm    = sum(precip, na.rm = TRUE),
    temp_mean_C  = mean(temp2_mavg, na.rm = TRUE),
    .groups      = "drop"
  )

ggplot(monthly_precip, aes(month, precip_mm)) +
  geom_col(fill = "#4575b4") +
  labs(
    x = NULL,
    y = "Precipitation (mm)",
    title = "Monthly precipitation — Konza Prairie 2024"
  ) +
  theme_bw()
```

## FW13 fire weather data

`ks_meso_fw13()` retrieves fire weather records in the standard FW13 format used by USDA Forest Service fire behavior modeling tools.

```{r}
#| label: fire-weather
#| eval: false

cache_file <- file.path(cache_dir, "ks_meso_fw13_konza_2024-04.rds")
if (!file.exists(cache_file)) {
  fw13_records <- ks_meso_fw13(
    station    = "Konza Prairie",
    start_date = "2024-04-01",
    end_date   = "2024-04-30"
  )
  saveRDS(fw13_records, cache_file)
} else {
  fw13_records <- readRDS(cache_file)
}

head(fw13_records, 3)
```
