---
title: "Physical properties and unit conversions"
vignette: >
  %\VignetteIndexEntry{Physical properties and unit conversions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

preMetabolizer includes small physical-property and unit-conversion helpers for
common hydrology and atmospheric science tasks. These functions return plain
numeric vectors, making them easy to use in base R, dplyr pipelines, and model
input tables.

```{r}
#| label: libraries
#| message: false

library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Flow conversions

`convert_flow()` converts discharge between cubic feet per second (`"cfs"`),
cubic meters per second (`"cms"`), and liters per second (`"lps"`).

```{r}
#| label: flow-conversions

flow <- tibble::tibble(discharge_cfs = c(1, 5, 25, 100)) |>
  mutate(
    discharge_cms = convert_flow(discharge_cfs, from = "cfs", to = "cms"),
    discharge_lps = convert_flow(discharge_cfs, from = "cfs", to = "lps")
  )

flow
```

## Pressure conversions and elevation correction

`convert_pressure()` converts barometric pressure between common units.
`correct_bp()` adjusts pressure from a weather station elevation to a site
elevation.

```{r}
#| label: pressure

convert_pressure(101.325, from = "kPa", to = "atm")
convert_pressure(1013.25, from = "hPa", to = "Pa")

correct_bp(
  station_bp = 101.3,
  air_temp = 15,
  station_elev = 300,
  site_elev = 500,
  from_units = "kPa",
  to_units = "kPa"
)
```

Elevation correction is useful when a nearby weather station is not at the same
elevation as a stream site.

```{r}
#| label: pressure-table

site_pressure <- tibble::tibble(
  station_bp_kPa = c(101.3, 100.9, 100.5),
  air_temp_C = c(15, 17, 18),
  station_elev_m = 300,
  site_elev_m = 500
) |>
  mutate(
    site_bp_kPa = correct_bp(
      station_bp = station_bp_kPa,
      air_temp = air_temp_C,
      station_elev = station_elev_m,
      site_elev = site_elev_m
    )
  )

site_pressure
```

## Water density and water height

`calc_water_density()` calculates pure-water density from temperature.
`calc_water_height()` converts pressure-sensor readings into water height.

```{r}
#| label: water-density

temps <- tibble::tibble(water_temp = seq(0, 30, by = 5)) |>
  mutate(density_kg_m3 = calc_water_density(water_temp))

temps
```

For vented sensors, `sensor_kPa` is already the pressure from the water column.
For unvented sensors, provide atmospheric pressure so it can be subtracted from
the absolute sensor pressure.

```{r}
#| label: water-height

water_level <- tibble::tibble(
  sensor_kPa = c(18.9, 19.2, 19.5),
  atmo_kPa = c(101.1, 101.0, 100.9),
  water_temp = c(14.8, 15.0, 15.2)
) |>
  mutate(
    vented_height_m = calc_water_height(
      sensor_kPa = sensor_kPa,
      water_temp = water_temp,
      type = "vented"
    ),
    unvented_height_m = calc_water_height(
      sensor_kPa = sensor_kPa + atmo_kPa,
      atmo_kPa = atmo_kPa,
      water_temp = water_temp,
      type = "unvented"
    )
  )

water_level
```

## Modeled light

`calc_light()` estimates photosynthetically active radiation (PAR) from solar
time and site coordinates. Input times should be mean solar time.

```{r}
#| label: modeled-light

utc <- seq(
  as.POSIXct("2024-06-21 00:00:00", tz = "UTC"),
  as.POSIXct("2024-06-21 23:00:00", tz = "UTC"),
  by = "hour"
)

light <- tibble::tibble(
  dateTime_UTC = utc,
  solar_time = convert_UTC_to_solartime(utc, longitude = -96.6)
) |>
  mutate(
    light_PAR = calc_light(
      solar.time = solar_time,
      latitude = 39.1,
      longitude = -96.6
    )
  )

head(light)
```

```{r}
#| label: plot-light

ggplot(light, aes(dateTime_UTC, light_PAR)) +
  geom_line(color = "#2c7fb8") +
  labs(
    x = NULL,
    y = "Modeled PAR"
  ) +
  theme_bw()
```
