---
title: "Working with TexMesonet data"
vignette: >
  %\VignetteIndexEntry{Working with TexMesonet data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

[TexMesonet](https://www.texmesonet.org/) is a statewide earth observation
network managed by the Texas Water Development Board (TWDB). TWDB stations
collect near-real-time weather and soil observations, including air
temperature, humidity, precipitation, wind, solar radiation, soil temperature,
and soil moisture.

preMetabolizer provides three helpers for the public TexMesonet API:

- `tex_meso_stations()` retrieves TWDB station metadata.
- `tex_meso_current()` retrieves the most recent observation from each TWDB
  station.
- `tex_meso_timeseries()` retrieves recent time-series observations for one
  station.

> **Note:** These functions contact the TexMesonet API and require an internet
> connection. Code chunks marked with `#| eval: false` will not run during
> package installation. Run them interactively in your own session.

```{r}
#| label: libraries
#| message: false

library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Discover TWDB stations

Use `tex_meso_stations()` to retrieve station names, IDs, display IDs,
coordinates, elevation, activity status, and online dates.

```{r}
#| label: stations
#| eval: false

stations <- tex_meso_stations()

glimpse(stations)
```

The `active` and `displayed` arguments make it easy to focus on stations that
are currently operating and shown by TexMesonet.

```{r}
#| label: active-stations
#| eval: false

active_stations <- tex_meso_stations(active = TRUE, displayed = TRUE)

active_stations |>
  select(stationId, stationName, displayId, county, latitude, longitude) |>
  arrange(county, stationName)
```

For station time-series requests, keep the integer `stationId`. This example
finds stations in Blanco County and selects one station ID for later use.

```{r}
#| label: choose-station
#| eval: false

blanco_stations <- active_stations |>
  filter(county == "Blanco") |>
  select(stationId, stationName, displayId, latitude, longitude, elevation)

blanco_stations

site_id <- blanco_stations$stationId[1]
```

## Retrieve current observations

`tex_meso_current()` returns the most recent available observation from each
TWDB station. Not every station measures every parameter, so some columns may
contain missing values.

```{r}
#| label: current
#| eval: false

current <- tex_meso_current()

glimpse(current)
```

The TexMesonet API reports units in a separate object. preMetabolizer stores
that object as a data-frame attribute.

```{r}
#| label: current-units
#| eval: false

attr(current, "units")
```

For a quick station check, join current observations to station metadata or
filter directly by `stationId`.

```{r}
#| label: current-station
#| eval: false

current |>
  filter(stationId == site_id) |>
  select(
    stationId,
    name,
    recordedTime,
    airTemp,
    humidity,
    precip,
    windSpeed,
    airPressure
  )
```

## Retrieve recent time series

`tex_meso_timeseries()` retrieves observations for a single station over a
look-back window measured in minutes. By default, `variable = "all"` uses the
TexMesonet charting-fields endpoint and returns all fields available for that
station.

```{r}
#| label: all-timeseries
#| eval: false

recent <- tex_meso_timeseries(
  site_id = site_id,
  prior_minutes = 24 * 60
)

glimpse(recent)
```

The returned `dateTime` column is parsed as UTC.

```{r}
#| label: datetime
#| eval: false

range(recent$dateTime, na.rm = TRUE)
attr(recent$dateTime, "tzone")
```

## Retrieve one variable

TexMesonet also provides smaller single-variable endpoints. Use the `variable`
argument when you only need one series.

```{r}
#| label: temperature-series
#| eval: false

temperature <- tex_meso_timeseries(
  site_id = site_id,
  prior_minutes = 24 * 60,
  variable = "temperature"
)

temperature |>
  arrange(dateTime) |>
  tail()

attr(temperature, "units")
```

Available single-variable values are `"temperature"`, `"humidity"`,
`"barometric_pressure"`, `"precip"`, and `"wind_speed"`.

## Example workflow: station weather summary

The charting-fields endpoint is useful when you want to prepare local
meteorological covariates for a stream site. The example below summarizes
hourly temperature, humidity, precipitation, and wind speed from the recent
station data.

```{r}
#| label: hourly-summary
#| eval: false

hourly_weather <- recent |>
  mutate(hour = lubridate::floor_date(dateTime, "hour")) |>
  group_by(hour) |>
  summarise(
    air_temp_C = mean(airTemp, na.rm = TRUE),
    humidity_pct = mean(humidity, na.rm = TRUE),
    precip_mm = sum(precip, na.rm = TRUE),
    wind_speed_m_s = mean(windSpeed, na.rm = TRUE),
    .groups = "drop"
  )

hourly_weather
```

Plotting the recent series can help identify gaps or station behavior before
combining meteorological data with stream logger observations.

```{r}
#| label: plot-temperature
#| eval: false

ggplot(hourly_weather, aes(hour, air_temp_C)) +
  geom_line(color = "#2c7fb8") +
  geom_point(color = "#2c7fb8", size = 1.5) +
  labs(
    x = NULL,
    y = "Air temperature (°C)",
    title = "Recent TexMesonet air temperature"
  ) +
  theme_bw()
```

## API scope

These functions use the lightweight TexMesonet API endpoints for TWDB stations.
For historical custom downloads, longer date ranges, or non-TWDB provider
networks shown in the TexMesonet map viewer, use the TexMesonet Custom
Downloads page or the source network recommended by TWDB.
