---
title: "Working with TexMesonet data"
vignette: >
  %\VignetteIndexEntry{Working with TexMesonet data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

cache_dir <- tools::R_user_dir("preMetabolizer", which = "cache")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

.run <- tryCatch({
  httr2::request("https://www.texmesonet.org") |>
    httr2::req_timeout(5) |>
    httr2::req_perform()
  TRUE
}, error = function(e) {
  length(list.files(cache_dir, pattern = "^tex_meso_")) > 0
})

knitr::opts_chunk$set(eval = .run)
```

## Introduction

[TexMesonet](https://www.texmesonet.org/) is a statewide earth observation network managed by the Texas Water Development Board (TWDB). TWDB stations collect near-real-time weather and soil observations, including air temperature, humidity, precipitation, wind, solar radiation, soil temperature, and soil moisture.

preMetabolizer provides three helpers for the public TexMesonet API:

- `tex_meso_stations()` retrieves TWDB station metadata.
- `tex_meso_current()` retrieves the most recent observation from each TWDB station.
- `tex_meso_time_series()` retrieves recent time-series observations for one station.

> **Note:** These functions contact the TexMesonet API and require an internet connection. Code chunks that call the API will not run during package installation if the service is unreachable. Run them interactively in your own session.

```{r}
#| label: libraries
#| message: false

library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Caching downloaded data

Repeated calls to the TexMesonet API download the same data on every run. This vignette saves each result to a local cache directory the first time it is downloaded and reloads from disk on subsequent runs. The cache lives in `tools::R_user_dir("preMetabolizer", which = "cache")`, a platform-appropriate, user-specific directory that persists across sessions. Each data-fetching chunk below checks for a cached `.rds` file, downloads and saves on the first run, and reloads from disk on all subsequent runs.

## Discover TWDB stations

Use `tex_meso_stations()` to retrieve station names, IDs, display IDs, coordinates, elevation, activity status, and online dates.

```{r}
#| label: stations

cache_file <- file.path(cache_dir, "tex_meso_stations.rds")
if (!file.exists(cache_file)) {
  stations <- tex_meso_stations()
  saveRDS(stations, cache_file)
} else {
  stations <- readRDS(cache_file)
}

glimpse(stations)
```

The `active` and `displayed` arguments make it easy to focus on stations that are currently operating and shown by TexMesonet.

```{r}
#| label: active-stations

cache_file <- file.path(cache_dir, "tex_meso_stations_active.rds")
if (!file.exists(cache_file)) {
  active_stations <- tex_meso_stations(active = TRUE, displayed = TRUE)
  saveRDS(active_stations, cache_file)
} else {
  active_stations <- readRDS(cache_file)
}

active_stations |>
  select(station_id, station_name, display_id, county, latitude, longitude) |>
  arrange(county, station_name)
```

For station time-series requests, keep the integer `station_id`. This example finds stations in Blanco County and selects one station ID for later use.

```{r}
#| label: choose-station

blanco_stations <- active_stations |>
  filter(county == "Blanco") |>
  select(station_id, station_name, display_id, latitude, longitude, elevation)

blanco_stations

site_id <- blanco_stations$station_id[1]
```

## Retrieve current observations

`tex_meso_current()` returns the most recent available observation from each TWDB station. Not every station measures every parameter, so some columns may contain missing values.

```{r}
#| label: current

cache_file <- file.path(cache_dir, "tex_meso_current.rds")
if (!file.exists(cache_file)) {
  current <- tex_meso_current()
  saveRDS(current, cache_file)
} else {
  current <- readRDS(cache_file)
}

glimpse(current)
```

The TexMesonet API reports units in a separate object. preMetabolizer stores that object as a data-frame attribute.

```{r}
#| label: current-units

attr(current, "units")
```

For a quick station check, join current observations to station metadata or filter directly by `station_id`.

```{r}
#| label: current-station

current |>
  filter(station_id == site_id) |>
  select(
    station_id,
    station_name,
    recorded_time,
    air_temp,
    humidity,
    precip,
    wind_speed,
    air_pressure
  )
```

## Retrieve recent time series

`tex_meso_time_series()` retrieves observations for a single station over a look-back window measured in minutes. By default, `variable = "all"` uses the TexMesonet charting-fields endpoint and returns all fields available for that station.

```{r}
#| label: all-timeseries

cache_file <- file.path(cache_dir, "tex_meso_time_series_blanco.rds")
if (!file.exists(cache_file)) {
  recent <- tex_meso_time_series(
    site_id = site_id,
    prior_minutes = 24 * 60
  )
  saveRDS(recent, cache_file)
} else {
  recent <- readRDS(cache_file)
}

glimpse(recent)
```

The returned `date_time` column is parsed as UTC.

```{r}
#| label: datetime

range(recent$date_time, na.rm = TRUE)
attr(recent$date_time, "tzone")
```

## Retrieve one variable

TexMesonet also provides smaller single-variable endpoints. Use the `variable` argument when you only need one series.

```{r}
#| label: temperature-series

cache_file <- file.path(cache_dir, "tex_meso_time_series_blanco_temperature.rds")
if (!file.exists(cache_file)) {
  temperature <- tex_meso_time_series(
    site_id = site_id,
    prior_minutes = 24 * 60,
    variable = "temperature"
  )
  saveRDS(temperature, cache_file)
} else {
  temperature <- readRDS(cache_file)
}

temperature |>
  arrange(date_time) |>
  tail()

attr(temperature, "units")
```

Available single-variable values are `"temperature"`, `"humidity"`, `"barometric_pressure"`, `"precip"`, and `"wind_speed"`.

## Example workflow: station weather summary

The charting-fields endpoint is useful when you want to prepare local meteorological covariates for a stream site. The example below summarizes hourly temperature, humidity, precipitation, and wind speed from the recent station data.

```{r}
#| label: hourly-summary

hourly_weather <- recent |>
  mutate(hour = lubridate::floor_date(date_time, "hour")) |>
  group_by(hour) |>
  summarise(
    air_temp_C = mean(air_temp, na.rm = TRUE),
    humidity_pct = mean(humidity, na.rm = TRUE),
    precip_mm = sum(precip, na.rm = TRUE),
    wind_speed_m_s = mean(wind_speed, na.rm = TRUE),
    .groups = "drop"
  )

hourly_weather
```

Plotting the recent series can help identify gaps or station behavior before combining meteorological data with stream logger observations.

```{r}
#| label: plot-temperature

ggplot(hourly_weather, aes(hour, air_temp_C)) +
  geom_line(color = "#2c7fb8") +
  geom_point(color = "#2c7fb8", size = 1.5) +
  labs(
    x = NULL,
    y = "Air temperature (°C)",
    title = "Recent TexMesonet air temperature"
  ) +
  theme_bw()
```

## API scope

These functions use the lightweight TexMesonet API endpoints for TWDB stations. For historical custom downloads, longer date ranges, or non-TWDB provider networks shown in the TexMesonet map viewer, use the TexMesonet Custom Downloads page or the source network recommended by TWDB.
