---
title: "Time-series and summary utilities"
vignette: >
  %\VignetteIndexEntry{Time-series and summary utilities}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

preMetabolizer includes utilities for regularizing logger data, converting between UTC and solar time, assigning seasons, flagging outliers, and summarizing variability or flow distributions.

```{r}
#| label: libraries
#| message: false

library(preMetabolizer)
library(dplyr)
library(ggplot2)
```

## Align to even timesteps

`even_timesteps()` snaps logger data onto an evenly spaced, rounded time grid. The time step is inferred from the data, timestamps are rounded to that step (here, hourly readings starting at 00:59 become 01:00, 02:00, ...), and values are linearly interpolated onto the grid. Grid times inside data gaps are left `NA`.

```{r}
#| label: even-timesteps

logger <- tibble::tibble(
  DateTime_UTC = as.POSIXct(
    c(
      "2024-06-01 00:59:00",
      "2024-06-01 01:59:00",
      "2024-06-01 03:59:00"
    ),
    tz = "UTC"
  ),
  temp_water = c(18.1, 18.0, 17.8)
)

even_timesteps(logger)
```

For multi-site data, provide the site column so each site is aligned independently.

```{r}
#| label: even-timesteps-site

multi_site <- tibble::tibble(
  Site = c("A", "A", "A", "B", "B"),
  DateTime_UTC = as.POSIXct(
    c(
      "2024-06-01 00:00:00",
      "2024-06-01 01:00:00",
      "2024-06-01 03:00:00",
      "2024-06-01 00:00:00",
      "2024-06-01 00:30:00"
    ),
    tz = "UTC"
  )
)

even_timesteps(multi_site, site_col = "Site")
```

## Convert UTC and solar time

`convert_to_solar_time()` and `convert_from_solar_time()` move between UTC and local solar time at a site longitude. Mean solar time is the time basis expected by stream metabolism models. Use `type = "apparent"` to additionally apply the equation of time (via `SunCalcMeeus::solar_time()`).

```{r}
#| label: solar-time

utc <- as.POSIXct("2024-06-21 18:00:00", tz = "UTC")

solar <- convert_to_solar_time(utc, longitude = -96.6)

solar

convert_from_solar_time(solar, longitude = -96.6)
```

## Assign seasons

`get_season()` classifies dates into astronomical seasons using the actual equinox and solstice dates for each year (computed via Meeus's algorithms). It accepts a `hemisphere` argument and returns an ordered factor.

```{r}
#| label: seasons

dates <- as.Date(c(
  "2024-01-15",
  "2024-04-15",
  "2024-07-15",
  "2024-10-15"
))

tibble::tibble(
  date = dates,
  season = get_season(dates)
)
```

## Flag potential outliers

`flag_z()` applies a moving-window robust Z-score. The default return is a character flag vector. Set `return_z = TRUE` when you also want the Z-scores.

```{r}
#| label: flag-z

temperature <- c(18.1, 18.2, 18.0, 18.3, 29.9, 18.4, 18.2)

flag_z(temperature, width = 5)

flag_z(temperature, width = 5, return_z = TRUE)
```

## Summary statistics

`calc_cv()` calculates the coefficient of variation. Use `robust = TRUE` to summarize relative variability with the median and MAD instead of mean and standard deviation.

```{r}
#| label: cv-mode

discharge <- c(0.12, 0.18, 0.15, 1.4, 0.09)

calc_cv(discharge)
calc_cv(discharge, robust = TRUE)

calc_mode(c("riffle", "run", "riffle", "pool", "run"), multi = "all")
```

## Histogram bin widths

`calc_bin_width()` implements several common histogram rules. This is helpful when plotting distributions that should use a consistent, data-driven bin width.

```{r}
#| label: bin-width

set.seed(1)
values <- stats::rexp(200)

calc_bin_width(values, method = "fd")
calc_bin_width(values, method = "doane")
```

```{r}
#| label: plot-histogram

ggplot(tibble::tibble(values = values), aes(values)) +
  geom_histogram(
    binwidth = calc_bin_width(values, method = "fd"),
    color = "white",
    fill = "#2c7fb8"
  ) +
  labs(x = "Value", y = "Count") +
  theme_bw()
```

## Flow exceedance probabilities

`calc_exceedance_prob()` ranks flows with the Weibull plotting-position formula. Higher flows receive lower exceedance probabilities. The C++ implementation, `rcpp_calc_exceedance_prob()`, returns the same shape and is useful for large vectors.

```{r}
#| label: exceedance

flows <- tibble::tibble(
  flow_cms = c(10, 5, 0, 15, 8, NA, 0, 20)
) |>
  mutate(
    exceedance = calc_exceedance_prob(flow_cms),
    exceedance_no_zero = rcpp_calc_exceedance_prob(flow_cms, remove_zeros = TRUE)
  )

flows
```

```{r}
#| label: plot-exceedance

flows |>
  filter(!is.na(exceedance_no_zero)) |>
  ggplot(aes(exceedance_no_zero, flow_cms)) +
  geom_point(color = "#2c7fb8", size = 2) +
  scale_x_continuous(labels = \(x) paste0(round(100 * x), "%")) +
  labs(
    x = "Exceedance probability",
    y = "Flow"
  ) +
  theme_bw()
```
